---
title: "UMAP"
#author: "Elisa"
date: "May 27, 2023"
output: 
  html_document:
    toc: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to UMAP (Uniform Manifold Approximation and Projection)
UMAP is, like t-SNE, a dimensionality reduction technique used for visualizing and analyzing high-dimensional data. 
UMAP computes pairwise distances between data points (usually Euclidean distance), constructs a graph representation, and optimizes a low-dimensional embedding that reflects the data structure using a student's t-distribution probability function.

UMAP uses a number-of-nearest-neighbors parameter in the computation of similarity scores, which are then used in the embedding to reflect the data structure.
The number-of-nearest-neighbors parameter is also the most impactful parameter to adjust.

t-SNE and UMAP are quite similar methods.

## Setup
```{r, warning=FALSE, message = FALSE}
library(tidyverse)
library(Seurat)
library(anndata)
library(pdfCluster)
library(gplots)
```


```{r load pilot set}
setwd("~/sc_covid_PiB2023/data_science_project/Data")
data <- read_h5ad("Pilot_2_rule_them_ALL.h5ad") # our pilot data
```


## First PCA, followed by UMAP
In this section we run UMAP on the data on which PCA has been used to dimension reduced.

### UMAP for every celletype
We start by running it on the different cell types using 'RunUMAP' from the 'Seurat' package.
Running UMAP on a cell type level allows for the comparison of infected and uninfected cells within each cell type.
```{r class.source = 'fold-hide', warning=FALSE, message=FALSE}
# Making a dataframe (used for plotting w. ggplots) with UMAP coordinates
celltypes <- unique(data$obs$cellType)
df <- as.data.frame(data$obs) %>%
  add_column(UMAP_1 = 0) %>%
  add_column(UMAP_2 = 0)

for (c in celltypes) {
  c_subset = data[which(data$obs$cellType == c)] # Subsetting based on celltype
  PCA <-
    RunPCA(
      c_subset$X,
      assay = NULL,
      npcs = 50,
      rev.pca = F,
      weight.by.var = T,
      verbose = F
    )
  c_cells_UMAP <- RunUMAP(
    PCA[],
    umap.method = "uwot",
    assay = "RNA",
    seed.use = 1,
    verbose = F
  )

df[df$cellType == c, ]$UMAP_1 <- c_cells_UMAP[[, 1]]
df[df$cellType == c, ]$UMAP_2 <- c_cells_UMAP[[, 2]]
}
```

#### Infection status:
Plot colored by infection status

```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = Is_infected)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"))    +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Viral load:
Plot colored by viral load
```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = viralLoad)) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Patient ID
Plot colored by patient id

```{r class.source = 'fold-hide'}
df %>% 
  arrange(viralLoad) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = PatientID)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

### Clustering on cell type
We try clustering for infection status on cell level.

```{r class.source = 'fold-hide', warning=FALSE}

celltypes <- unique(data$obs$cellType)

matrix_of_ari <- matrix(1:48, nrow = 8, ncol = 6,
                      dimnames = list(celltypes,
                                      c("PCA_kmeans", "PCA_hier", "scVI_kmeans", "scVI_hier", "cor_kmeans", "cor_hier")))

n_clusters = 2 # number of clusters

df["kmeans"] <- 0
df["hierarchical"] <- 0

for (c in celltypes){ 
  c_subset = df[df$cellType == c,]
  max_ari = -1
  
# Kmeans:
  cluster <- kmeans(c_subset[, 9:10], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  matrix_of_ari[c, "PCA_kmeans"]  = adj.rand.index(cluster$cluster, c_subset$Is_infected) # ARI for kmeans clustering
  df[df$cellType == c,]$kmeans <- factor(cluster$cluster)
  
# Hierarchical
  for (m in c("complete", "single", "average", "centroid")){
  dm <- dist(as.data.frame(c_subset[, 9:10])) # Distance matrix
  hc <- hclust(dm, method = "centroid") # simple dendrogram
  clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
  ari = adj.rand.index(clusterCutS, c_subset$Is_infected)
  df[df$cellType == c,]$hierarchical <- factor(clusterCutS)
  if (max_ari < ari) {
      matrix_of_ari[c, "PCA_hier"] <- ari
      max_ari = ari}
  }}
matrix_of_ari <- round(matrix_of_ari, 4)
matrix_of_ari[,1:2]
```

#### Kmeans

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(kmeans))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Kmeans on UMAP", subtitle = "Dimension reduction with PCA\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

#### Hierarchical

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(hierarchical))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Hierarchical on UMAP", subtitle = "Dimension reduction with PCA\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

### Loop To Optimize Model Parameters with Adjusted Rand Index
This loop-based approach utilizes the adjusted Rand index (ARI) as a measure to guide the search for the best parameter values. By systematically iterating through different combinations of parameters and evaluating their performance using the ARI, this approach helps identify the parameter values that yield the highest similarity between predicted and true labels. It saves the coordinates in the dataset to be used for clustering plots in 'Clean_Clustering'.

```{r, eval = FALSE}
dist_values <- c(0.15, 0.25, 0.5, 0.8, 0.99)
near <- c(5, 10, 25, 50, 70, 100)

n_clusters = 8 # number of clusters
max_ARI_kmeans = 0
max_ARI_hier = 0

for (k in near) {
  for (c in dist_values) {
    umap_test <- RunUMAP(
      as.matrix(data$obsm$X_pca),
      umap.method = "uwot",
      assay = "RNA",
      seed.use = 1,
      n.neighbors = k,
      min.dist = c,
      spread = 1
      #verbose = TRUE
      )
  cluster <- kmeans(umap_test[[, 1:2]], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  ARI_kmeans = adj.rand.index(cluster$cluster, data$obs$cellType) # ARI for kmeans clustering

  if (max_ARI_kmeans < ARI_kmeans) {
    max_ARI_kmeans = ARI_kmeans
    best_kmeans <- umap_test}
  
  for (m in c("complete", "single", "average", "centroid")){
    dm <- dist(as.data.frame(umap_test[[, 1:2]])) # Distance matrix
    hc <- hclust(dm, method = m) # simple dendrogram
    clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
    ARI_hier = adj.rand.index(clusterCutS, data$obs$cellType)
    
    if (max_ARI_hier < ARI_hier) {
      max_ARI_hier = ARI_hier
      best_hier <- umap_test}
    }
  }
}
# We save the coordinates that gives the highest ARI in the pilot set to plot the clusterings later.
if (max_ARI_kmeans > max_ARI_hier) {
  matrix_data <-
    matrix(
      data = c(best_kmeans[[, 1]], best_kmeans[[, 2]]),
      nrow = length(best_kmeans[[, 1]]),
      ncol = 2
    )
  data$obsm$X_PCA_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
} else {
  matrix_data <-
    matrix(
      data = c(best_hier[[, 1]], best_hier[[, 2]]),
      nrow = length(best_hier[[, 1]]),
      ncol = 2
    )
  data$obsm$X_PCA_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
}
head(data$obsm$X_PCA_UMAP)
```


## First scVI, followed by UMAP:
In this section we run UMAP on the data on which scVI has been used to dimension reduced.

### UMAP for every celletype
We start by running it on the different cell types using 'RunUMAP' from the 'Seurat' package.
Running UMAP on a cell type level allows for the comparison of infected and uninfected cells within each cell type.
```{r class.source = 'fold-hide', warning=FALSE, message=FALSE}
celltypes <- unique(data$obs$cellType)
df <- as.data.frame(data$obs) %>% 
  add_column(UMAP_1 = 0) %>% 
  add_column(UMAP_2 = 0)

for (c in celltypes){ 
  
  c_subset = data[which(data$obs$cellType == c)]
  c_cells_UMAP <- RunUMAP(
    c_subset$obsm$X_scVI,
    umap.method = "uwot",
    assay = "RNA",
    seed.use = 1,
    verbose = F
  )

df[df$cellType == c, ]$UMAP_1 <- c_cells_UMAP[[, 1]]
df[df$cellType == c, ]$UMAP_2 <- c_cells_UMAP[[, 2]]
}
```

#### Infection status:
Plots colored by the infection status of each cell
```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = Is_infected)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"))    +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Viral load:
Plots colored by the viral load of each cell
```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = viralLoad)) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Patient ID
Plots colored by patient id
```{r class.source = 'fold-hide'}
df %>% 
  arrange(viralLoad) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = PatientID)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

### Clustering for each celltype
We try clustering for infection status on cell level.

```{r, warning=FALSE, message=FALSE}

celltypes <- unique(data$obs$cellType)

n_clusters = 2 # number of clusters

df["kmeans"] <- 0
df["hierarchical"] <- 0

for (c in celltypes){ 
  c_subset = df[df$cellType == c,]
  max_ari = -1
  
# Kmeans:
  cluster <- kmeans(c_subset[, 9:10], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  matrix_of_ari[c, "scVI_kmeans"] = adj.rand.index(cluster$cluster, c_subset$Is_infected) # ARI for kmeans clustering
  df[df$cellType == c,]$kmeans <- factor(cluster$cluster)
  
# Hierarchical
  for (m in c("complete", "single", "average", "centroid")){
  dm <- dist(as.data.frame(c_subset[, 9:10])) # Distance matrix
  hc <- hclust(dm, method = "centroid") # simple dendrogram
  clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
  ari = adj.rand.index(clusterCutS, c_subset$Is_infected)
  df[df$cellType == c,]$hierarchical <- factor(clusterCutS)
  if (max_ari < ari) {
      matrix_of_ari[c, "scVI_hier"] <- ari
      max_ari = ari}
  }}
matrix_of_ari <- round(matrix_of_ari, 4)
matrix_of_ari[,3:4]
```

#### Kmeans

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(kmeans))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Kmeans on UMAP", subtitle = "Dimension reduction with scVI\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

#### Hierarchical

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(hierarchical))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Hierarchical on UMAP", subtitle = "Dimension reduction with scVI\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

### Loop To Optimize Model Parameters with Adjusted Rand Index
This loop-based approach utilizes the adjusted Rand index (ARI) as a measure to guide the search for the best parameter values. By systematically iterating through different combinations of parameters and evaluating their performance using the ARI, this approach helps identify the parameter values that yield the highest similarity between predicted and true labels. It saves the coordinates in the dataset to be used for clustering plots in 'Clean_Clustering'.

```{r loop for umap parameters, message = FALSE, eval=FALSE}
dist_values <- c(0.15, 0.25, 0.5, 0.8, 0.99)
near <- c(5, 10, 25, 50, 70, 100)

n_clusters = 8 # number of clusters
max_ARI_kmeans = 0
max_ARI_hier = 0

for (k in near) {
  for (c in dist_values) {
    umap_test <- RunUMAP(
      data$obsm$X_scVI,
      umap.method = "uwot",
      assay = "RNA",
      seed.use = 1,
      n.neighbors = k,
      min.dist = c,
      spread = 1
      #verbose = TRUE
      )
  cluster <- kmeans(umap_test[[, 1:2]], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  ARI_kmeans = adj.rand.index(cluster$cluster, data$obs$cellType) # ARI for kmeans clustering

  if (max_ARI_kmeans < ARI_kmeans) {
    max_ARI_kmeans = ARI_kmeans
    kmeans_c = c
    kmeans_k = k
    best_kmeans <- umap_test}
  
  for (m in c("complete", "single", "average", "centroid")){
    dm <- dist(as.data.frame(umap_test[[, 1:2]])) # Distance matrix
    hc <- hclust(dm, method = m) # simple dendrogram
    clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
    ARI_hier = adj.rand.index(clusterCutS, data$obs$cellType)
    
    if (max_ARI_hier < ARI_hier) {
      best_linkage = m
      max_ARI_hier = ARI_hier
      hier_c = c
      hier_k = k
      best_hc = hc
      best_hier <- umap_test}
    }
  }
}
# We save the coordinates that gives the highest ARI in the pilot set to plot the clusterings later.
if (max_ARI_kmeans > max_ARI_hier) {
  matrix_data <-
    matrix(
      data = c(best_kmeans[[, 1]], best_kmeans[[, 2]]),
      nrow = length(best_kmeans[[, 1]]),
      ncol = 2
    )
  data$obsm$X_scVI_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
} else {
  matrix_data <-
    matrix(
      data = c(best_hier[[, 1]], best_hier[[, 2]]),
      nrow = length(best_hier[[, 1]]),
      ncol = 2
    )
  data$obsm$X_scVI_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
}
```


### Loop for heatmap purposes
This loop explores various parameter combinations to find the optimal configuration for hierarchical clustering, based on the highest Adjusted Rand Index (ARI) score. The ARI scores are stored in a dataframe for creating a heatmap which offers a visual representation of the explored parameter space.

```{r SUPER loop of only hierarchical clustering for heatmap purposes, warning=FALSE, message=FALSE}

df_ari <- data.frame(matrix(nrow = 6, ncol = 5))

dist_values <- c(0.15, 0.25, 0.5, 0.8, 0.99)
near <- c(5, 10, 25, 50, 70, 100)

colnames(df_ari) <- dist_values
rownames(df_ari) <- near

n_clusters = 8 # number of clusters

for (i in 1:nrow(df_ari)) {
  for (j in 1:ncol(df_ari)) {
    umap_test <- RunUMAP(
      data$obsm$X_scVI,
      umap.method = "uwot",
      assay = "RNA",
      seed.use = 1,
      n.neighbors = near[i],
      min.dist = dist_values[j],
      spread = 1,
      verbose = FALSE
      )
  dm <- dist(as.data.frame(umap_test[[, 1:2]])) # Distance matrix
  hc <- hclust(dm, method = "centroid") # simple dendrogram
  clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
  ARI <- adj.rand.index(clusterCutS, data$obs$cellType)
  df_ari[i, j] <- round(ARI, 4)
  }
}
```

#### Heatmap of ARI scores
The heatmap displays the ARI value of differnet combinations of parameters. The best score is marked with a blue border.

```{r class.source = 'fold-hide', warning=FALSE}
ari_matrix <- as.matrix(df_ari)

# Find the maximum value and its coordinates
max_coords <- as.vector(which(ari_matrix == max(ari_matrix), arr.ind = TRUE))



# Function to mark the max value
makeRects <- function(cells){
  xl=cells[2]-0.49
  yb=nrow(ari_matrix)+1-cells[1]-0.49
  xr=cells[2]+0.49
  yt=nrow(ari_matrix)+1-cells[1]+0.49
  rect(xl,yb,xr,yt,border="blue",lwd=3)
}


heatmap.2(
  ari_matrix,
  Colv = NA,
  Rowv = NA,
  dendrogram = "none",
  trace = "none",
  xlab = "min.dist",
  ylab = "n.neigbours",
  main = "Heatmap of ARI",
  notecol = "black",
  key = TRUE,
  cellnote = df_ari,
  add.expr = {makeRects(max_coords)})

```


## First scVI that has been batch corrected, followed by UMAP:
In this section we run UMAP on the data on which scVI with batch corrections has been used to dimension reduced.
The idea is that by applying batch correction with scVI, it becomes easier to identify and characterize cell types more accurately, enhancing downstream analyses such as clustering. By correcting we might also enhance the ability to discern true biological differences from technical impressions.

### UMAP for every cell type
We start by running it on the different cell types using 'RunUMAP' from the 'Seurat' package.
Running UMAP on a cell type level allows for the comparison of infected and uninfected cells within each cell type.
```{r class.source = 'fold-hide', warning=FALSE, message=FALSE}
celltypes <- unique(data$obs$cellType)
df <- as.data.frame(data$obs) %>% 
  add_column(UMAP_1 = 0) %>% 
  add_column(UMAP_2 = 0)

for (c in celltypes){ 
  
  c_subset = data[which(data$obs$cellType == c)]
  c_cells_UMAP <- RunUMAP(
    c_subset$obsm$X_scVI_corrected,
    umap.method = "uwot",
    assay = "RNA",
    seed.use = 1,
    verbose = F
  )

df[df$cellType == c, ]$UMAP_1 <- c_cells_UMAP[[, 1]]
df[df$cellType == c, ]$UMAP_2 <- c_cells_UMAP[[, 2]]
}
```

#### Infection status:
Plots colored by the infection status of each cell
```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = Is_infected)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  scale_color_manual(values = c("0" = "blue", "1" = "red"))    +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Viral load:
Plots colored by the viral load of each cell
```{r class.source = 'fold-hide'}
df %>%
  arrange(viralLoad) %>%
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = viralLoad)) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_color_gradient(low = "blue", high = "red") +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") +
  xlab("UMAP_1") +
  ylab("UMAP_2")
```

#### Patient ID
Plots colored by patient id
```{r class.source = 'fold-hide'}
df %>% 
  arrange(viralLoad) %>% 
  ggplot(aes(x = UMAP_1, y = UMAP_2, color = PatientID)) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "UMAP cell type") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

### Clustering for each celltype
To run this loop one must have already run the 'tSNE for every cell type' chunk

```{r, warning=FALSE, message=FALSE}

celltypes <- unique(data$obs$cellType)

n_clusters = 2 # number of clusters

df["kmeans"] <- 0
df["hierarchical"] <- 0

for (c in celltypes){ 
  c_subset = df[df$cellType == c,]
  max_ari=-1
  
# Kmeans:
  cluster <- kmeans(c_subset[, 9:10], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  matrix_of_ari[c, "cor_kmeans"] = adj.rand.index(cluster$cluster, c_subset$Is_infected) # ARI for kmeans clustering
  df[df$cellType == c,]$kmeans <- factor(cluster$cluster)
  
# Hierarchical
  for (m in c("complete", "single", "average", "centroid")){
  dm <- dist(as.data.frame(c_subset[, 9:10])) # Distance matrix
  hc <- hclust(dm, method = "centroid") # simple dendrogram
  clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
  ari = adj.rand.index(clusterCutS, c_subset$Is_infected)
  df[df$cellType == c,]$hierarchical <- factor(clusterCutS)
  if (max_ari < ari) {
      matrix_of_ari[c, "cor_hier"] <- ari
      max_ari = ari}
  }}
matrix_of_ari <- round(matrix_of_ari, 4)
matrix_of_ari[,5:6]

```

#### Kmeans

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(kmeans))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Kmeans on UMAP", subtitle = "Dimension reduction with scVI with batch correction\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

#### Hierarchical

```{r class.source = 'fold-hide'}
ggplot(df, aes(x = UMAP_1, y = UMAP_2 , color = factor(hierarchical))) +
  geom_point(size = 1.5, alpha = 0.6) +
  facet_wrap("cellType", scales = "free", labeller = labeller()) +
  theme(axis.text = element_blank(),
  axis.ticks = element_blank()) +
  labs(title = "Hierarchical on UMAP", subtitle = "Dimension reduction with scVI with batch correction\nClustered by infection; k = 2") + 
  xlab("UMAP_1") +
  ylab("UMAP_2") +
  theme(legend.position = "none")
```

A heatmap displaying the ARI score of each combination of dimention reduction methods and clustering method with the different celltypes.

```{r class.source = 'fold-hide'}
heatmap.2(
  matrix_of_ari,
  Colv = NA,
  Rowv = NA,
  dendrogram = "none",
  cexRow=0.8,
  cexCol = 0.8,
  trace = "none",
  main = "ARI with different methods",
  #col = colorRampPalette(c("white", "blue")), 
  notecol = "black",
  key = TRUE,
  cellnote = matrix_of_ari)
```

### Loop To Optimize Model Parameters with Adjusted Rand Index
This loop-based approach utilizes the adjusted Rand index (ARI) as a measure to guide the search for the best parameter values. By systematically iterating through different combinations of parameters and evaluating their performance using the ARI, this approach helps identify the parameter values that yield the highest similarity between predicted and true labels. It saves the coordinates in the dataset to be used for clustering plots in 'Clean_Clustering'.

```{r, eval = FALSE}
dist_values <- c(0.15, 0.25, 0.5, 0.8, 0.99)
near <- c(5, 10, 25, 50, 70, 100)

n_clusters = 8 # number of clusters
max_ARI_kmeans = 0
max_ARI_hier = 0

for (k in near) {
  for (c in dist_values) {
    umap_test <- RunUMAP(
      data$obsm$X_scVI_corrected ,
      umap.method = "uwot",
      assay = "RNA",
      seed.use = 1,
      n.neighbors = k,
      min.dist = c,
      spread = 1
      #verbose = TRUE
      )
  cluster <- kmeans(umap_test[[, 1:2]], iter = 10000, nstart = 100, n_clusters) # kmeans clustering
  ARI_kmeans = adj.rand.index(cluster$cluster, data$obs$cellType) # ARI for kmeans clustering

  if (max_ARI_kmeans < ARI_kmeans) {
    max_ARI_kmeans = ARI_kmeans
    best_kmeans <- umap_test}
  
  for (m in c("complete", "single", "average", "centroid")){
    dm <- dist(as.data.frame(umap_test[[, 1:2]])) # Distance matrix
    hc <- hclust(dm, method = m) # simple dendrogram
    clusterCutS <- cutree(hc, n_clusters) # data, number of clusters
    ARI_hier = adj.rand.index(clusterCutS, data$obs$cellType)
    
    if (max_ARI_hier < ARI_hier) {
      max_ARI_hier = ARI_hier
      best_hier <- umap_test}
    }
  }
}
# We save the coordinates that gives the highest ARI in the pilot set to plot the clusterings later.
if (max_ARI_kmeans > max_ARI_hier) {
  matrix_data <-
    matrix(
      data = c(best_kmeans[[, 1]], best_kmeans[[, 2]]),
      nrow = length(best_kmeans[[, 1]]),
      ncol = 2
    )
  data$obsm$X_scVI_corrected_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
} else {
  matrix_data <-
    matrix(
      data = c(best_hier[[, 1]], best_hier[[, 2]]),
      nrow = length(best_hier[[, 1]]),
      ncol = 2
    )
  data$obsm$X_scVI_corrected_UMAP <-  matrix_data
  write_h5ad(data, '../Data/Pilot_2_rule_them_ALL.h5ad')
}
```